The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching

### 2016-09-29 article.tex
smore more edits

@@

We here outline various new and improved APIs in the CDK library since the two previous
publications.

@@

  Atom type perception is a core cheminformations functionality: the atom types describe
  chemical features of atoms, such as the number of neighbors, possible formal charges,
  (approximate) hybridization, electron distribution over orbitals, etc. However, the
  CDK commonly integrated atom type perception into algorithms, resulting in multiple
  and diverging copies of similar atom type schemes. With the need to add many
  new atom types for previously uncovered element, but also charged and radical atom
  types, a new approach was needed.

  This CDK version has isolated a new atom typing framework, removing the perception of
  atom types from various algorithms, allowing the perception to be tested separately.
  The new code defines the atom types using a list that specifies for each type
  the element symbol, hybridization, formal charge, number of lone pairs, and
  an enumeration of the bond orders (see Figure~1). This list of properties captures the
  information needed for the various algorithms in the CDK. For example,
  hybridization information can be used in certain aromaticity models (see later),
  and the lone pair information is needed for resonance structure calculation
  needed, for example, for Gasteiger $\pi$-charges.

  A reference perception class, CDKAtomTypeMatcher, has been written that perceives these atom types, and
  validates the perception automatically against the properties defined by the ontology.
  This class handles a variety in types of missing information, as commonly resulting
  from various (file) formats; for example, it can handle undefined hydrogen counts
  and undefined double bond positions if hybridization information is provided instead.
  That makes the perception code flexible but also more complex. Alternative
  algorithms to do the perception have not been explored. This reference
  implementation can be used on a single atom:

@@

  The preferred representation of stereochemistry is now for it to be stored at the molecule
  level as a \texttt{\textbf{StereoElement}}. In abstract terms a stereo element describes local
  geometry using a type, focus, carriers, and configuration (Figure~2\ref{fig:stereodatastructure}).
  Currently the most common types of stereochemistry are supported: Tetrahedral, Cis-trans isomerism 
  around a double bond, and Extended Tetrahedral. Rarer types of stereochemistry, such as: Square 
  Planar, Trigonal Bipyramidal, Octahedral, could easily be incorporated into the chosen description 
  given sufficient demand from the community.

>>>>

In this section we described the specifics of new APIs and
improvements to pre-existing methods that are available in the latest
CDK.  We then discuss how we have improved and formalized the
development model for the project using unit testing, code review and
guidelines for handling version control. Finally we report on the
availability of binary distributions of the library, allowing users to
include specific modules (and their dependencies) of the CDK in their
own projects (as opposed to developers who work on the CDK library
itself).

@@

We here outline various new and improved APIs in the CDK library since the two previous
publications (corresponding to version XXX)

@@

  Atom type perception is a core cheminformations functionality: the
  atom types describe chemical features of atoms, such as the number
  of neighbors, possible formal charges, (approximate) hybridization,
  electron distribution over orbitals and so on. However, previous
  versions of the CDK implemented atom type perception as part of
  different algorithms, resulting in duplicated and sometimes
  divergent typing schemes. As a result it was cumbersome to add new
  atom types and implement support for new charged and radical species
  in a consistent manner.

  This CDK version has a new, centralized atom typing framework,
  removing the perception of atom types from various algorithms. This
  allows for a consistent and extensive typing scheme, that can be
  also be tested independently of other code.  The new code defines
  the atom types using a list that specifies for each type the element
  symbol, hybridization, formal charge, number of lone pairs, and an
  enumeration of the bond orders (see Figure~\ref{fig:atomtype}). This list of
  properties captures the information needed for the various
  algorithms in the CDK. For example, hybridization information can be
  used in certain aromaticity models (see later), and the lone pair
  information is needed for resonance structure calculation needed,
  for example, for Gasteiger $\pi$-charges.

  A reference implementation, \texttt{CDKAtomTypeMatcher}, has been
  written that perceives these atom types, and validates the
  perception automatically against the properties defined by the
  ontology.  This class handles a variety in types of missing
  information, as commonly resulting from various (file) formats; for
  example, it can handle undefined hydrogen counts and undefined
  double bond positions if hybridization information is provided
  instead.  That makes the perception code flexible but also more
  complex. Alternative algorithms for atom typing have not been
  explored. This reference implementation can be used on a single
  atom:

@@

  The preferred representation of stereochemistry is now for it to be stored at the molecule
  level as a \texttt{\textbf{StereoElement}}. In abstract terms a stereo element describes local
  geometry using a type, focus, carriers, and configuration (Figure~\ref{fig:stereodatastructure}).
  Currently the most common types of stereochemistry are supported: Tetrahedral, Cis-trans isomerism 
  around a double bond, and Extended Tetrahedral. Rarer types of stereochemistry, such as: Square 
  Planar, Trigonal Bipyramidal, Octahedral, could easily be incorporated into the chosen description 
  given sufficient demand from the community.
